% Appendix 4
\glsresetall % reset the glossary to expand acronyms again
\chapter{Example Programs}\label{ch:ExamplePrograms}

The following are a series of example programs written in the Garter programming
language. The majority of these example programs were adapted from the series of
example programs to be shown to teachers interested in using the Turing
programming for education.

\section{Hello World}

This program demonstrates a minimal program in Garter. It shows the lack of
boilerplate in basic Garter programs, and demonstrates the basic use of
the \code{print} statement.

\lstinputlisting[language=Python]{figs/helloworld.py}

\section{Fibonacci}

This program demonstrates function definitions in Garter, recursion support
and the if statement.

\lstinputlisting[language=Python]{figs/fibonacci.py}

\section{Adder}

This program demonstrates user input through the \code{input} function-like
expression, variable declaration, type coersions, and the \code{while} loop.

\lstinputlisting[language=Python]{figs/adder.py}

\section{Sorter}

This program demonstrates user input through the \code{input} function-like
expression, lists, typed varaible declarations, and the \code{for} loop.

\lstinputlisting[language=Python]{figs/sorter.py}

\section{Linked List}

This program demonstrates a basic implementation of a linked-list structure,
along with some basic implementation methods. It demonstrates class
declarations, and method declarations.

\lstinputlisting[language=Python]{figs/linkedlist.py}

\section{Boxes}

This program demonstrates basic input, string manipulations, and the use of the
\code{range} looping construct.

\lstinputlisting[language=Python]{figs/boxes.py}

\section{Expression Parser}

This program demonstrates the implementation of a trivial expression parser.

\lstinputlisting[language=Python]{figs/expn.py}

\section{Hanoi}

This program demonstrates an implementation of the tower of hanoi algorithm.

\lstinputlisting[language=Python]{figs/hanoi.py}

\section{Random Number Generation}

This program demonstrates the generation of a random numbers using Garter.

\lstinputlisting[language=Python]{figs/rand.py}

